---
title: "Survey of Employment, Payrolls and Hours"
author: "Trevor Tombe"
date: "January 2019"
output:
  html_document:
    includes:
      after_body: footer.html
    theme: spacelab
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
always_allow_html: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(message=F, warning=F,fig.width = 8,fig.height = 4.5,fig.align="center",cache=TRUE,echo=F)

```

```{r load data, include=F}
source("core.R")

# Fetches Latest SEPH Data
seph<-getTABLE("14100223")
seph<-seph %>% rename(NAICS=North.American.Industry.Classification.System..NAICS.)
sourcetable<-"Statistics Canada table 14-10-0223"
today=max(seph$Ref_Date) # the latest month in the data

# Constructs Employment Data
sephemp<-seph %>%
  filter(Estimate=="Employment for all employees",
         NAICS=="Industrial aggregate including unclassified businesses")

# Wage and Salary Data
wagedata<-getTABLE("14100222")

thismonth=as.character(today)
lastmonth=as.character(today-1/12)
lastyear=as.character(today-1)

```

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Each month Statistics Canada releases comprehensive data on the state of Canada's labour markets. The Survey of Employment, Payroll and Hours is based on firm-level administrative payroll data. This is distinct from the Labour Force Survey, which is another important source of labour market information.

This page summarizes all of the latest results from that data.
</div>

# Change in Employment, by Province {.tabset .tabset-fade}
Displays the change in payroll employment and employment rates. First, since the start of Alberta's recession. Then compares provinces.

## Alberta

```{r growth}
startdate="Oct 2014"
plotdata<-sephemp %>%
  filter(Ref_Date>=startdate & GEO=="Alberta") %>%
  group_by(GEO) %>%
  mutate(change=Value-max(Value*(Ref_Date==startdate)))
ggplot(plotdata,aes(Ref_Date,change/1000))+
  geom_line(size=2) +
  geom_point(data=plotdata %>% filter(Ref_Date==max(Ref_Date)),size=4)+
	geom_hline(aes(yintercept=0), colour="black", size=1) +
  geom_text(data=plotdata[plotdata$Ref_Date==max(plotdata$Ref_Date),],
            aes(x=Ref_Date,y=change/1000,label=paste(round(change/1000),"k",sep="")),
            hjust=0,vjust=-0.5,fontface="bold",size=5,color="firebrick") +
  mytheme+
  scale_y_continuous(breaks=pretty_breaks(n=8)) +
  scale_x_yearmon(limit=c(2014.75,max(plotdata$Ref_Date)+1/12),breaks=pretty_breaks(n=8),format = "%b\n%Y") +
	labs(x="",y="Thousands of Jobs",title=paste("Change in",plotdata$GEO[1],"Payroll Jobs Since",startdate),
		subtitle="Note: The 2015-16 recession started in October 2014. Source: Statistics Canada table 14-10-0223",caption="Graph by @trevortombe")

province<-"Alberta"
lfsdata<-fromJSON("http://economicdashboard.alberta.ca/Download/DownloadFile?extension=JSON&requestUrl=http%3A%2F%2Feconomicdashboard.alberta.ca%2Fapi%2FEmployment")
startdate="Dec 2014"
plotdata <- lfsdata %>%
  mutate(Ref_Date=as.yearmon(as.Date(When))) %>%
  filter(Ref_Date>=startdate,Characteristic=="Employment") %>%
  select(Ref_Date,emp=Alberta) %>%
  mutate(change=(emp-emp[1]),
         Group="Labour Force Survey ('Employment')") %>%
  bind_rows(plotdata<-sephemp %>%
              filter(Ref_Date>=startdate & GEO=="Alberta") %>%
              group_by(GEO) %>%
              mutate(change=Value-max(Value*(Ref_Date==startdate))) %>% 
              mutate(Group="Payroll Data ('Jobs')") %>% select(Ref_Date,change,Group))
ggplot(plotdata,aes(Ref_Date,change/1000,group=Group,color=Group))+
  geom_hline(yintercept=0,size=1) +
  geom_line(size=2,lineend="round") +
	mytheme+
	scale_colour_brewer(name="",palette="Set1") +
	scale_y_continuous(breaks=pretty_breaks(n=8)) +
  scale_x_yearmon(breaks=pretty_breaks(n=8),format = "%b\n%Y") +
  labs(x="",y="Thousands of Persons",title=paste("Change in Alberta Employment Since",startdate),
		subtitle=paste("Source: Own calculations from Statistics Canada data table 14-10-0287 and",sourcetable),caption="Graph by @trevortombe")

```

```{r level in Alberta}
startdate<-"Jan 2012"
plotdata<-sephemp %>%
  filter(Ref_Date>=startdate & GEO=="Alberta")
ggplot(plotdata,aes(Ref_Date,Value/1000))+
  annotate("rect",xmin=as.numeric(as.yearmon("Oct 2014")), xmax=as.numeric(as.yearmon("Oct 2016")), 
           ymin=-Inf, ymax=+Inf, alpha=0.2, fill="dodgerblue")+
  annotate("text",x=2015.71, y=Inf, vjust=1,
           label="Recession",color="black",size=3,fontface="italic")+
  geom_line(size=2,color="firebrick") +
  geom_point(data=plotdata %>% filter(Ref_Date==max(Ref_Date)),size=3,color="firebrick")+
  mytheme+
  scale_y_continuous(breaks=pretty_breaks(n=8),label=comma) +
  scale_x_yearmon(breaks=pretty_breaks(n=8),format = "%Y") +
  labs(x="",y="Thousands of Jobs",title=paste("Total Payroll Employment in",plotdata$GEO),
       subtitle=paste("Source: CANSIM",sourcetable),caption="Graph by @trevortombe")
```

## All Provinces
```{r all provinces}
startdate="Oct 2014"
plotdata<-sephemp %>%
  filter(Ref_Date>=startdate & GEO!="Yukon" & GEO!="Northwest Territories" & GEO!="Canada" & GEO!="Nunavut") %>%
  group_by(GEO) %>%
  mutate(change=Value/Value[1]-1)
ggplot() +
	geom_line(size=1,show.legend=FALSE) +
	geom_line(data=plotdata[(plotdata$GEO!="Alberta" & plotdata$GEO!="Canada"),], aes(x=Ref_Date, y=change,group=GEO), size = 1, color="gray") +
	geom_line(data=plotdata[plotdata$GEO=="British Columbia",], aes(x=Ref_Date, y=change), size = 2, color="royalblue4") +
	geom_line(data=plotdata[plotdata$GEO=="Alberta",], aes(x=Ref_Date, y=change), size = 2, color="firebrick") +
  geom_line(data=plotdata[plotdata$GEO=="Saskatchewan",], aes(x=Ref_Date, y=change), size = 2, color="darkgreen") +
  geom_hline(aes(yintercept=0), colour="black", size=1) +
  mytheme+
  scale_y_continuous(breaks=pretty_breaks(n=8),label=percent) +
  scale_x_yearmon(breaks=pretty_breaks(n=8),format = "%b\n%Y") +
  labs(x="",y="Per Cent Change (%)",title=paste("Change in Payroll Jobs in Canada,",startdate,"to",today),
		subtitle=paste("Source: CANSIM",sourcetable),caption="Graph by @trevortombe") +
	annotate(geom="text", x=2016.25, y=.07, label="British Columbia",color="royalblue4",size=3,fontface="bold") +
	annotate(geom="text", x=2016.75, y=-.07, label="Alberta",color="firebrick",size=3,fontface="bold") +
  annotate(geom="text", x=2016.5, y=-.025, label="Saskatchewan",color="darkgreen",size=3,fontface="bold") +
  annotate(geom="text", x=2015.25, y=.04, label="All Others",color="grey60",size=3,fontface="bold")
```

# By Sector
Growth in employment, by sector.

```{r by sector}
plotdata<-as_tibble(seph) %>%
  filter(GEO=="Alberta" & Ref_Date>="Jan 2016" & 
           !(NAICS %in% c("Industrial aggregate including unclassified businesses",
                          "Industrial aggregate excluding unclassified businesses",
                          "Goods producing industries","Service producing industries",
                          "Durable goods","Non-durable goods","Trade"))) %>%
  select(Ref_Date,Estimate,NAICS,Value) %>%
  mutate(
    NAICS=replace(NAICS,NAICS=="Administrative and support, waste management and remediation services","Admin Support & Waste"),
    NAICS=replace(NAICS,NAICS=="Other services (except public administration)","Other Service"),
    NAICS=replace(NAICS,NAICS=="Mining, quarrying, and oil and gas extraction","Mining/Oil/Gas"),
    NAICS=replace(NAICS,NAICS=="Forestry, logging and support","Forestry"),
    NAICS=replace(NAICS,NAICS=="Accommodation and food services","Hotel/Restaurants"),
    NAICS=replace(NAICS,NAICS=="Arts, entertainment and recreation","Arts, enter., rec."),
    NAICS=replace(NAICS,NAICS=="Professional, scientific and technical services","Professional/Scientific"),
    NAICS=replace(NAICS,NAICS=="Real estate and rental and leasing","Real Estate"),
    NAICS=replace(NAICS,NAICS=="Transportation and warehousing","Transport/Storage"),
    NAICS=replace(NAICS,NAICS=="Information and cultural industries","Information/Culture"),
    NAICS=replace(NAICS,NAICS=="Health care and social assistance","Health/Social"),
    NAICS=replace(NAICS,NAICS=="Management of companies and enterprises","Mgmt of Companies")
    ) %>%
  arrange(Estimate,NAICS,Ref_Date) %>%
  group_by(Estimate,NAICS) %>%
  mutate(change=Value-lag(Value,12))
ggplot(filter(plotdata,Estimate=="Employment for all employees" & Ref_Date==max(Ref_Date)),
       aes(reorder(NAICS,change),change)) +
  geom_bar(stat="identity",fill="dodgerblue3") +
  geom_hline(yintercept=0,size=1)+
  coord_flip() +
  mythemebarflip+
  scale_y_continuous(breaks=pretty_breaks(n=6),label=comma)+
  labs(x="",y="Number of Payroll Jobs",
       title=paste("Change in Alberta's Payroll Jobs,",lastyear,"to",thismonth),
       subtitle="Source: Statistics Canada data table 14-10-0223.",caption="Graph by @trevortombe.")

plotdata<-as_tibble(seph) %>%
  filter(GEO=="Alberta" & Ref_Date>="Jan 2016" & 
           !(NAICS %in% c("Industrial aggregate including unclassified businesses",
                          "Industrial aggregate excluding unclassified businesses",
                          "Goods producing industries","Service producing industries",
                          "Durable goods","Non-durable goods","Trade"))) %>%
  select(Ref_Date,Estimate,NAICS,Value) %>%
  mutate(
    NAICS=replace(NAICS,NAICS=="Administrative and support, waste management and remediation services","Admin Support & Waste"),
    NAICS=replace(NAICS,NAICS=="Other services (except public administration)","Other Service"),
    NAICS=replace(NAICS,NAICS=="Mining, quarrying, and oil and gas extraction","Mining/Oil/Gas"),
    NAICS=replace(NAICS,NAICS=="Forestry, logging and support","Forestry"),
    NAICS=replace(NAICS,NAICS=="Accommodation and food services","Hotel/Restaurants"),
    NAICS=replace(NAICS,NAICS=="Arts, entertainment and recreation","Arts, enter., rec."),
    NAICS=replace(NAICS,NAICS=="Professional, scientific and technical services","Professional/Scientific"),
    NAICS=replace(NAICS,NAICS=="Real estate and rental and leasing","Real Estate"),
    NAICS=replace(NAICS,NAICS=="Transportation and warehousing","Transport/Storage"),
    NAICS=replace(NAICS,NAICS=="Information and cultural industries","Information/Culture"),
    NAICS=replace(NAICS,NAICS=="Health care and social assistance","Health/Social"),
    NAICS=replace(NAICS,NAICS=="Management of companies and enterprises","Mgmt of Companies")
    ) %>%
  arrange(Estimate,NAICS,Ref_Date) %>%
  group_by(Estimate,NAICS) %>%
  mutate(change=Value-lag(Value,1))
ggplot(filter(plotdata,Estimate=="Employment for all employees" & Ref_Date==max(Ref_Date)),
       aes(reorder(NAICS,change),change)) +
  geom_bar(stat="identity",fill="dodgerblue3") +
  geom_hline(yintercept=0,size=1)+
  coord_flip() +
  mythemebarflip+
  scale_y_continuous(breaks=pretty_breaks(n=6),label=comma)+
  labs(x="",y="Number of Payroll Jobs",
       title=paste("Change in Alberta's Payroll Jobs,",lastmonth,"to",thismonth),
       subtitle="Source: Statistics Canada data table 14-10-0223.",caption="Graph by @trevortombe.")

# Shift in Alberta's employment composition
start="Oct 2014"
seph<-seph %>% data.table()
plotdata<-seph[GEO=="Alberta" & Estimate=="Employment for all employees" & Ref_Date>=start & NAICS!="Industrial aggregate including unclassified businesses" &
                 NAICS!="Industrial aggregate excluding unclassified businesses" & NAICS!="Durable goods" & NAICS!="Trade" &
                 NAICS!="Goods producing industries" & NAICS!="Service producing industries" & NAICS!="Non-durable goods"]
plotdata[,totalemp:=sum(Value),by=Ref_Date]
plotdata[,share:=Value/totalemp]
plotdata<-plotdata[Ref_Date==start | Ref_Date==max(Ref_Date)]

test2<-dcast(plotdata[,.(Ref_Date,NAICS,share)],NAICS~Ref_Date)
test2[NAICS=="Administrative and support, waste management and remediation services"]$NAICS<-"Admin Support & Waste"
test2[NAICS=="Other services (except public administration)"]$NAICS<-"Other Service"
test2[NAICS=="Mining, quarrying, and oil and gas extraction"]$NAICS<-"Mining/Oil/Gas"
test2[NAICS=="Forestry, logging and support"]$NAICS<-"Forestry"
test2[NAICS=="Accommodation and food services"]$NAICS<-"Hotel/Restaurants"
test2[NAICS=="Arts, entertainment and recreation"]$NAICS<-"Arts, enter., rec."
test2[NAICS=="Professional, scientific and technical services"]$NAICS<-"Professional/Scientific"
test2[NAICS=="Real estate and rental and leasing"]$NAICS<-"Real Estate"
test2[NAICS=="Transportation and warehousing"]$NAICS<-"Transport/Storage"
test2[NAICS=="Information and cultural industries"]$NAICS<-"Information/Culture"
test2[NAICS=="Health care and social assistance"]$NAICS<-"Health/Social"
test2[NAICS=="Management of companies and enterprises"]$NAICS<-"Mgmt of Corps."
setnames(test2,start,"start")
setnames(test2,paste(today),"end")
test2<-test2[order(test2$start),]
test2$NAICS<-factor(test2$NAICS,levels=test2$NAICS)
test2[,change:=end-start]
test2[,arrowend:=change-0.001*(change>0)+0.001*(change<0)]
ggplot(test2, aes(x=start, xend=end, y=NAICS, group=NAICS)) + 
  geom_segment(aes(xend=start+arrowend,yend=NAICS),color="royalblue",
               arrow = arrow(type="closed",length = unit(0.15,"cm")))+
  geom_dumbbell(color="royalblue", 
                size=1, 
                size_x=2,
                colour_xend="firebrick",
                size_xend=4)+
  geom_text(aes(x=.05, y="Mining/Oil/Gas", label=paste(today)),
            color="firebrick", hjust=1, size=3, nudge_y=0.5)+
  geom_text(aes(x=0.068, y="Mining/Oil/Gas", label="Oct 2014"),
            color="royalblue", hjust=0, size=3, nudge_y=0.5)+
  geom_vline(xintercept = 0,size=1)+
  mytheme+
  scale_x_continuous(expand=c(0,0),limits=c(0,0.13),label=percent,breaks=pretty_breaks(n=8)) + 
  labs(x="Share of Total Payroll Employment (%)", 
       y=NULL, 
       title="Where Are Alberta's Jobs? The Changing Composition Since Oct 2014", 
       subtitle="Source: CANSIM 281-0063. October 2014 is used as the start of the recession.",
       caption="Graph by @trevortombe")
```

# Earnings

```{r earnings}
plotdata<-seph %>%
  filter(Estimate=="Average weekly earnings including overtime for all employees" & 
                 NAICS=="Industrial aggregate excluding unclassified businesses" & Ref_Date==max(Ref_Date)) %>%
  left_join(provnames,by="GEO") %>%
  filter(!(short %in% c("CAN","NU","NT","YT")))
ggplot(plotdata, aes(x=short,y=Value)) +
	geom_col(fill="dodgerblue3") +
  geom_text(aes(label=paste0("$",round(Value))),vjust=1.5,color="white",fontface="bold")+
	geom_hline(aes(yintercept=0), colour="black", size=1) +
	mythemebar +
  scale_y_continuous(limits=c(0,1400),breaks=pretty_breaks(n=8),labels=dollar,expand=c(0,0)) +
	labs(x="",y="Dollars per Week",title=paste0("Average Weekly Earnings, by Province (",thismonth,")"),
		subtitle=paste("Source: CANSIM",sourcetable),caption="Graph by @trevortombe")

startdate="Jan 2011"
plotdata<-seph %>%
  filter(Estimate=="Average weekly earnings including overtime for all employees" &
                  NAICS=="Industrial aggregate excluding unclassified businesses" & Ref_Date>=startdate &
           !(GEO %in% c("Nunavut","Northwest Territories","Yukon")))
ggplot(plotdata,aes(Ref_Date,Value,group=GEO))+
	geom_line(size = 1, color="gray") +
  geom_line(data=plotdata %>% filter(GEO %in% c("Alberta","Saskatchewan","Canada","Newfoundland and Labrador")),
            aes(Ref_Date,Value,group=GEO,color=GEO), size = 2) +
	mytheme +
	scale_colour_brewer(name="",palette="Set1") +
	scale_y_continuous(limit=c(700,1200),breaks=pretty_breaks(n=10),labels=dollar) +
	scale_x_yearmon(format = "%Y",n=8) +
	annotate(geom="text", x=2016, y=770,hjust=0, label="All Other Provinces",color="grey70",size=4,fontface="bold") +
	labs(x="",y="Dollars",title=paste("Average Weekly Earnings,",startdate,"to",today),
		subtitle=paste("Source: CANSIM",sourcetable),caption="Graph by @trevortombe")

ggplot(plotdata %>% filter(GEO=="Alberta"),aes(Ref_Date,Value))+
  annotate("rect",xmin=as.numeric(as.yearmon("Oct 2014")), xmax=as.numeric(as.yearmon("Oct 2016")), 
           ymin=-Inf, ymax=+Inf, alpha=0.2, fill="dodgerblue")+
  annotate("text",x=2015.71, y=Inf, vjust=1.25,label="Recession",color="black",size=3,fontface="italic")+
  geom_line(size=2,color="firebrick") +
  geom_point(data=plotdata %>% filter(GEO=="Alberta",Ref_Date==max(Ref_Date)),size=3,color="firebrick")+
  mytheme +
  scale_y_continuous(labels=dollar,expand=c(0,0),limit=c(1000,1200)) +
  scale_x_yearmon(format = "%Y",breaks=seq(2011,2018)) +
  labs(x="",y="Dollars",title=paste("Average Weekly Earnings in Alberta,",startdate,"to",today),
       subtitle=paste("Source: CANSIM",sourcetable),caption="Graph by @trevortombe")

plotdata<-seph %>%
  filter(Estimate=="Average weekly earnings including overtime for all employees" & GEO %in% c("Canada","Alberta") &
           NAICS=="Industrial aggregate excluding unclassified businesses") %>%
  group_by(GEO) %>%
  mutate(change=Value/lag(Value,12)-1) %>%
  filter(Ref_Date>="Jan 2012")
ggplot(plotdata,aes(Ref_Date,change,group=GEO,color=GEO))+
  annotate("rect",xmin=as.numeric(as.yearmon("Oct 2014")), xmax=as.numeric(as.yearmon("Oct 2016")), 
           ymin=-Inf, ymax=+Inf, alpha=0.2, fill="dodgerblue")+
  annotate("text",x=2015.71, y=max(pretty(plotdata$change)), hjust=0.5, 
           label="AB Recession",color="black",size=3,fontface="italic")+
  geom_line(size=2) +
	geom_hline(aes(yintercept=0), colour="black", size=1) +
	mytheme +
  theme(legend.position = c(0.9,0.25))+
	scale_colour_brewer(name="",labels=c("Alberta","Canada"),palette="Set1") +
	scale_y_continuous(breaks=pretty_breaks(n=10),labels=percent) +
	scale_x_yearmon(format = "%b\n%Y",n=8) +
	labs(x="",y="Year-over-Year % Change",title="Change in Average Weekly Earnings (YoY %)",
		subtitle=paste("Source: CANSIM",sourcetable),caption="Graph by @trevortombe")
```

## Hourly Wages vs Salaries

```{r wages}
## Weekly Earnings Change, hourly vs salary
sourcetable="Statistics Canadad data table 14-10-0222"
province="Alberta"
startdate="Oct 2014"
plotdata<-wagedata %>%
  filter(GEO==province & (Estimate=="Average weekly earnings including overtime for hourly employees" |
	   Estimate=="Average weekly earnings including overtime for salaried employees")) %>%
  group_by(Estimate) %>%
  mutate(YoY=Value/lag(Value,12)-1) %>%
  filter(Ref_Date>=startdate) %>%
  mutate(change=Value/Value[1]-1)
ggplot(plotdata,aes(Ref_Date,change,group=Estimate,color=Estimate))+
	geom_line(size=2) +
	geom_hline(aes(yintercept=0), colour="black", size=1) +
	mytheme +
	scale_colour_brewer(name="",palette="Set1",label=c("Hourly Employees","Salaried Employees")) +
  scale_y_continuous(breaks=pretty_breaks(n=10),labels=percent) +
	scale_x_yearmon(format = "%b\n%Y") +
	labs(x="",y="Per Cent Change",title=paste("Change in",province,"Weekly Earnings Since",startdate),
		subtitle=paste("Source: ",sourcetable,sep=""),caption="Graph by @trevortombe")
```

```{r hourly_emp}
# Hourly vs Salaried employment
plotdata<-wagedata %>%
  filter(GEO==province & Ref_Date>=startdate &
	(Estimate=="Employment for hourly paid employees" |
	   Estimate=="Employment for salaried employees")) %>%
  group_by(Estimate) %>%
  mutate(change=Value-Value[1])
ggplot(plotdata,aes(Ref_Date,change/1000,group=Estimate,color=Estimate))+
  geom_line(size=2) +
	geom_hline(aes(yintercept=0), colour="black", size=1) +
	mytheme +
  scale_colour_brewer(name="",palette="Set1",label=c("Hourly Employees","Salaried Employees")) +
  scale_y_continuous(breaks=pretty_breaks(n=8),labels=comma) +
	scale_x_yearmon(breaks=pretty_breaks(n=8),format = "%b\n%Y") +
	labs(x="",y="Thousands of Jobs",title=paste("Change in",province,"Payroll Employment Since",startdate),
	     subtitle=paste("Source: ",sourcetable,sep=""),caption="Graph by @trevortombe")
```

```{r wage_vs_hours}
plotdata<-wagedata %>%
  filter(GEO==province & Ref_Date>=startdate &
	(Estimate=="Average hourly earnings including overtime for hourly employees" |
	   Estimate=="Average weekly hours including overtime for hourly employees")) %>%
  group_by(Estimate) %>%
  mutate(change=Value/Value[1]-1)
ggplot(plotdata,aes(Ref_Date,change,group=Estimate,color=Estimate))+
  geom_line(size=2) +
	geom_hline(aes(yintercept=0), colour="black", size=1) +
	mytheme +
  scale_colour_brewer(name="",palette="Set1",label=c("Hourly Wages","Hours per Week")) +
  scale_y_continuous(breaks=pretty_breaks(n=8),labels=percent,limit=c(-0.08,0.06)) +
	scale_x_yearmon(format = "%Y") +
	labs(x="",y="Per Cent Change",title=paste("Change in",province,"Hours/Wages for Hourly Workers since",startdate),
	     subtitle=paste("Source: ",sourcetable,sep=""),caption="Graph by @trevortombe")
```

```{r hours}
plotdata<-wagedata %>%
  filter(GEO==province & Estimate=="Average weekly hours including overtime for hourly employees") %>%
  filter(Ref_Date>="Jan 2010")
ggplot(plotdata,aes(Ref_Date,Value))+
  annotate("rect",xmin=as.numeric(as.yearmon("Oct 2014")), xmax=as.numeric(as.yearmon("Oct 2016")), 
           ymin=-Inf, ymax=+Inf, alpha=0.2, fill="dodgerblue")+
  annotate("text",x=2015.71, y=Inf, vjust=1.25, 
           label="Recession",color="black",size=3,fontface="italic")+
  geom_line(size=2,color="firebrick")+
  mytheme +
  scale_y_continuous(breaks=pretty_breaks(n=6),labels=comma) +
  scale_x_yearmon(format="%Y")+
  labs(x="",y="Hours per Week",title="Average weekly hours in Alberta",
       subtitle=paste("Source: ",sourcetable,"-- including overtime; for hourly employees only."),caption="Graph by @trevortombe")
```


# Oil and Gas Detail

```{r oil and gas}
jobdetailed<-getTABLE("14100201")
plotdata<-jobdetailed %>%
  filter(GEO=="Alberta" & Type.of.employee=="All employees" &
           North.American.Industry.Classification.System..NAICS. %in% c("Oil and gas extraction",
                        "Mining and quarrying (except oil and gas)",
                        "Support activities for mining, and oil and gas extraction")) %>%
  rename(NAICS=North.American.Industry.Classification.System..NAICS.) %>%
  mutate(NAICS=replace(NAICS,NAICS=="Mining and quarrying (except oil and gas)","Mining and quarrying"),
         NAICS=replace(NAICS,NAICS=="Support activities for mining, and oil and gas extraction","Support activities"),
         Ref_Date=as.yearmon(Ref_Date,"%Y/%m"))
ggplot(plotdata %>% filter(Ref_Date>="Jan 2013"),aes(x=Ref_Date,y=Value,group=NAICS,color=NAICS)) +
  annotate("rect",xmin=as.numeric(as.yearmon("Oct 2014")), xmax=as.numeric(as.yearmon("Oct 2016")), 
           ymin=-Inf, ymax=+Inf, alpha=0.2, fill="dodgerblue")+
  annotate("text",x=2015.71, y=max(pretty(plotdata$Value)), hjust=0.5, 
           label="AB Recession",color="black",size=3,fontface="italic")+
  geom_line(size=2) +
	geom_hline(aes(yintercept=0), colour="black", size=1) +
	mytheme +
	scale_colour_brewer(name="",palette="Set1") +
	scale_y_continuous(breaks=pretty_breaks(n=10),labels=comma) +
	scale_x_yearmon(format="%Y") +
	labs(x="",y="Persons",title="Payroll Employment in Alberta Mining, Oil and Gas",
		subtitle="Source: CANSIM 281-0023",caption="Graph by @trevortombe")
```

# Diffusion Index

```{r diffusion}
test<-jobdetailed %>%
  filter(!(North.American.Industry.Classification.System..NAICS. %in% c("Industrial aggregate including unclassified businesses",
                                                                        "Industrial aggregate excluding unclassified businesses",
                                                                        "Goods producing industries","Manufacturing",
                                                                        "Non-durable goods","Durable goods",
                                                                        "Service producing industries","Trade")),
         Type.of.employee=="All employees",GEO=="Alberta") %>%
  group_by(GEO,North.American.Industry.Classification.System..NAICS.) %>%
  arrange(GEO,North.American.Industry.Classification.System..NAICS.) %>%
  mutate(grow=Value>lag(Value,12)) %>%
  filter(!is.na(grow)) %>%
  group_by(GEO,Ref_Date) %>%
  summarise(share=mean(grow)) %>%
  filter(Ref_Date>="Jan 2005")
ggplot(test,aes(Ref_Date,share))+geom_line(size=2,color="firebrick")+
  geom_hline(yintercept=0.5,size=1)+
  mytheme+
  scale_y_continuous(label=percent)+
  scale_x_continuous(breaks=pretty_breaks(n=6))+
  labs(x="",y="Per Cent",
       title="The Breadth of Alberta's Jobs Recovery (a `Diffusion Index`)",
       subtitle="Displays the share of disaggregated NAICS sectors with higher payroll employment than 12 months prior.
When above the black line, more sectors are expanding than shrinking. Below the line, more sectors are shrinking than expanding.
Source: Own calculations from Statistics Canada 14-10-0201. Note: I exclude high-level aggregates, but some double counting remains.",
       caption="Graph by @trevortombe")
```

# Vacancy-to-Unemployment

```{r vacancy_data}
vanunemp<-getTABLE("14100226")
```

```{r vacancy_plots}
plotdata<-vanunemp %>%
  filter(Job.vacancy.statistics=="Unemployment-to-job vacancies ratio",
         North.American.Industry.Classification.System..NAICS.=="All unemployed",
         GEO %in% c("Canada","Alberta","Saskatchewan"))
ggplot(plotdata,aes(Ref_Date,Value,group=GEO,color=GEO)) +
  annotate("rect",xmin=as.numeric(as.yearmon("Oct 2014")), xmax=as.numeric(as.yearmon("Oct 2016")), 
           ymin=0, ymax=+Inf, alpha=0.2, fill="dodgerblue")+
  annotate("text",x=2015.71, y=Inf, vjust=1, hjust=0.5, 
           label="AB Recession",color="black",size=3,fontface="italic")+
  geom_line(size=2) +
  geom_point(data=plotdata %>% filter(Ref_Date==max(Ref_Date)),size=4,show.legend = F)+
  geom_hline(aes(yintercept=0), colour="black", size=1) +
  geom_text(data=plotdata[plotdata$Ref_Date==max(vanunemp$Ref_Date),],
            aes(x=Ref_Date,y=Value,label=Value,color=GEO,group=GEO),nudge_x=0.25,vjust=-0.25,
            fontface="bold",size=4,show.legend=F) +
  mytheme +
  scale_colour_brewer(name="",palette="Set1") +
  scale_y_continuous(breaks=pretty_breaks(n=10),expand=c(0,0),limit=c(0,9)) +
  scale_x_yearmon(limit=c(NA,today+1/2),format = "%Y") +
  labs(x="",y="Workers per Vacancy",title="Unemployed Workers per Job Vacancy",
       subtitle="Source: Statistics Canada data table 14-10-0226",caption="Graph by @trevortombe")
plotdata<-vanunemp %>%
  filter(Job.vacancy.statistics=="Unemployment-to-job vacancies ratio",
         North.American.Industry.Classification.System..NAICS.=="All unemployed") %>%
  filter(Ref_Date==max(Ref_Date) | Ref_Date==(max(Ref_Date)-1)) %>%
  left_join(provnames,by="GEO") %>%
  filter(!(short %in% c("CAN","NU","YT","NT")) & !is.na(short))
ggplot(plotdata, aes(x=short,y=Value,group=as.character(Ref_Date),fill=as.character(Ref_Date))) +
  geom_col(position="dodge") +
  geom_hline(aes(yintercept=0), colour="black", size=1) +
  mythemebar+
  theme(legend.position = c(0.1,0.8))+
  scale_fill_brewer(name="",palette="Set1") +
  scale_y_continuous(breaks=pretty_breaks(n=10),expand=c(0,0)) +
  labs(x="",y="Unemployed per Vacancy",title="Unemployed Workers per Job Vacancy",
       subtitle="Source: Statistics Canada data table 14-10-0226",caption="Graph by @trevortombe")
```

